\emph{dfoliatR}: An R package for detection and analysis of insect defoliation signals in tree rings

### 2020-04-09 paper_elsevier/paper_elsevier.tex
writing progress [2020-04-09]

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={dfoliatR: An R package for detection and analysis of insect defoliation signals in tree rings},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc header

\usepackage{amsmath}

@@

Variation in the width and morphology of annual radial growth rings in
wood permits dating and quantification of past forest insect defoliator
outbreaks. Defoliation can be distinguished from climate- and other
disturbance-related influences by comparing ring-width or other
annually-resolved features in the wood of host species to non-host
species or to climate records. The effect of defoliation on radial
growth of trees has been recognized since the mid-1800s, but work prior
to the 1980s was not cross-dated (based on annually-resolved and --dated
measurements), not standardized (quantitatively controlled for tree
geometry and age effects), and suffered from inadequate extraction of
the defoliation-related variability from variability associated with
climate and other factors (Swetnam et al. 1985, Speer 2010, Lynch 2012).
These and other problems may be circumvented by appropriate use of
dendrochronological techniques. Quantitative methodology to infer forest
defoliator outbreaks from cross-dated tree-ring records was developed in
the 1980s by Swetnam et al. (1985) for developing western spruce budworm
chronologies (Swetnam and Lynch 1989, 1993). The methodology has since
been successfully applied to a wide range of defoliator species, most of
which are conifer herbivores, and has evolved in sophistication and
application to a wide range of ecosystem situations (Lynch 2012).

The main dendrochronological tool for inferring, dating, and quantifying
defoliator outbreaks from tree-ring records has been the software
routine OUTBREAK (Swetnam et al. 1985, Holmes and Swetnam 1986, Swetnam
and Lynch 1989). OUTBREAK computes indices (described later in detail)
of suppressed growth by subtracting a detrended and standardized climate
series (a ``control'' chronology, usually a site chronology developed
from non-host trees or a gridded climate data point series) from host
individual-tree detrended and standardized radial growth series after
the host and non-host series have been brought to a common variance. If
the host and non-host species respond similarly to climate (which can be
tested), the derived series retains variability that the host and
non-host series do not have in common, generally the insect signal and
noise (unexplained variability). The user defines a rule base specifying
the magnitude and duration that a period of indexed growth suppression
must meet or surpass for a period of suppressed growth to be inferred as
a defoliation event at the tree level.

Though powerful, OUTBREAK is outdated and increasingly difficult to use
in modern computing environments. It was written in Fortran V with
inherently severe restrictions, as RAM and disk space were limited at
that time (256 kb and 10 MB, respectively) and Fortran conventions
imposed very strict formatting, file naming, and output conventions.
Windows-based execution operates in a DOS window, Apple-based xxx in
xxx, and provides no graphical interface or capabilities. Barriers to
batch operation impose burdens for analyses of larger data sets. We
developed dfoliaR as an \texttt{R}- and \texttt{dplR}-based routine to
overcome these issues.

\texttt{dfoliatR} adds to a growing suite of dendrochronology packages
the \texttt{R} computing environment (R Core Team 2019). Stemming from
the \texttt{dplR} library (Bunn 2008) that enables \texttt{R} users to
read and write an array of tree-ring data formats, standardize ring
width series, build and evaluate chronologies, and perform quality
control (to name a few), one can now also measure ring widths from
scanned images of preparred samples (Lara et al. 2015, Shi and Xiang
2019), perform and check crossdating (Bunn 2010), and perform many
analytical tests (Zang and Biondi 2015, Jevšenak and Levanič 2018).
Tools for assessing stand dynamics and disturbance analyses are under
rapid development, with new packages for assessing growth and release
events (\texttt{TRADER}: Altman et al. 2014), metrics of growth
resilience (\texttt{pointRes}: Maaten-Theunissen et al. 2015), and fire
history (\texttt{burnr}: Malevich et al. 2018). The key objective of
\texttt{dfoliatR} is to provide tools to identify and analyze insect
defoliation and outbreak events by building on the methods employed by
OUTBREAK. It capitilizes on the robust software already available in
\texttt{R} by using \texttt{dplR} data formats for incoming tree-ring
series and providing output data formats embodied by the
\texttt{tidyverse} (Wickham et al. 2019) that include efficient data
manipulation (\texttt{dplyr}: Wickham et al. 2020) and graphics
(\texttt{ggplot2}: Wickham 2016).

In this paper, we describe the statistical methods employed by
\texttt{dfoliatR}, its availability, compare results to those produced
by OUTBREAK, and present an example analyses. Users need not have much
experience in \texttt{R} to replicate the analyses and graphics as
presented. The \texttt{R} code presented below is executable in an
\texttt{R} session once the required libraries are installed and loaded.
Support documentation in addition to this paper is provided within the
package via standard help menus (accessed by typing \texttt{?} before a
function name) and on the package website
(\url{https://chguiterman.github.io/dfoliatR/}), which includes
up-to-date vignettes that describe in detail the functionality of the
software. Code to create a preprint of this manuscript including the
\texttt{R} scripts is available from
\url{https://github.com/chguiterman/dfoliatR_paper}.

@@

Users can develop these data sets in software of their choosing, such as
\texttt{dplR} (Bunn 2008) or ARSTAN (Cook and Holmes 1996). It is
important that the host-tree data include only one tree-ring series per
tree. Both \texttt{dplR} and ARSTAN have options for averaging multiple
sample series into a tree-level series.

At the heart of \texttt{dfoliatR} lies two functions:
\texttt{defoliate\_trees()} and \texttt{outbreak()}. These identify
defoliation events on individual trees and then composite across
multiple trees to identify outbreak events.

@@

The \texttt{defoliate\_trees()} function is usually be the point of
entry to the dfoliatR library. It performs two processes, removing
climate-related growth signals from the host-tree series and then
identifying tree-level defoliation events. The climatic or
non-defoliation signals in each host-tree series are characterized by a
non-host chronology or climate reconstruction. \texttt{dfoiatR} removes
that non-defoliation signal by subtracting the non-host series from each
host-tree series, which generates a residual index. In OUTBREAK, this
residual index was termed the ``corrected'' index. We call it the
``growth suppression index'' (GSI). The GSI is calculated the same as in
OUTBREAK (following Swetnam et al. 1985, Swetnam and Lynch 1989) for
each host tree as \begin{align}
\textrm{GSI}_{i} = \textrm{H}_{i} - \left( \textrm{NH}_{i} - \overline{\textrm{NH}} \right) \frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}} \
\end{align} where H and NH are the host-tree series and the non-host
chronology, in year i, respectively. Only the common period between the
host-tree series and the non-host chronology are used in Equation 1. The
host and non-host chronologies are brought to common variance by scaling
the non-host chronology by its mean
(\(\overline{\textrm{NH}} \approx 1.0\)) and multiplying by the ratio of
host and non-host standard deviations
(\(\frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}}\)), which
approximates the variance of the host tree series.

Negative departures in the normalized GSI that surpass user-defined
thresholds in duration and magnitude are defined as \emph{defoliation
events}. As in OUTBREAK, magnitude is assessed on a single year within
the departure sequence. The default setting is -1.28 (in units of
standard deviation), which was previously determined to be
representative of WSBW effects (\textbf{citation?}). Duration is
assessed by examining sequences of negative GSI before and after the
year of maximum departure. Each defoliation event is allowed one
positive excursion on each side of the maximum departure year. Duration
is assessed across the entire sequence that includes up to two positive
excursions. The default duration is eight years, as is commonly used in
WSBW studies (\textbf{citation?}). Different species of defoliation
insects vary in the length of defoliation and the degree to which they
can suppress tree growth. Researchers can, and should, adjust the
duration and magnitude parameters accordingly and critically evaluate
the results.

@@

Defoliation of one or a few trees does not constitute an outbreak. To
determine when defoliation becomes an \emph{outbreak event},
\texttt{dfoliatR} composites the individual tree defoliation series into
a site-level chronology with the \texttt{outbreak()} function. Users
have options to define the number and/or the proportion of trees
required for an event to be considered an outbreak. Three parameters
control the whether a defoliation event constitutes an outbreak: the
minimum number of trees available, the minimum number of trees recording
defoliation, and the percent of trees recording defoliation. The first
allows the researcher to make a judgement call as to the confidence
ascribed to reduced sample depth toward the ends of their chronologies,
thus compensating for the ``fading record problem'' (\textbf{Swetnam and
Fritts?}). The second two parameters adjust the scale of defoliation
considered to be an outbreak. Absolute numbers of trees and percentages
can be applied separately or in conjunction, following filtering
conventions in tree ring fire history studies (Malevich et al. 2018). We
urge users to carefully consider the choice of absolute numbers in
situations where the number of trees represented in the series varies
with time, or the choice of percentages when sample size is small.

@@

We tested \texttt{dfoliatR} against OUTBREAK by comparing GSI to
corrected indices for individual trees and years, outbreak status for
individual trees and years, and percentage of trees recording outbreaks
at the site level, using raw ring-width data from 8 sites in British
Columbia, Colorado, and New Mexico and author-provided non-host site
chronologies.

We detrended host data for both dfoliatR and OUTBREAK using ARSTAN 6.1
(\textbf{downloaded 21 April 2002 from dpl; xxx Ann check the date, as
21 April 2002 is 6.05P}) with default double detrending (128 year
wavelength and a 50\% smoothing spline) and event thresholds of -1.28
normalized index and 8 years. Attempts to detrend with ARSTAN 44 in
\texttt{dplR} were problematic for two reasons. Some versions of ARSTAN,
including ARSTAN 44, do not include an option for producing tree-level
averages from multiple cores from the same tree (.tre files). More
importantly, ring-width indices from ARSTAN 44 truncated the process at
either end when sample size fell below 5 xxx (Jodi, what was the
minimum?), and produced different values for years near the truncation
point than did dpl-ARSTAN 6.05P. The majority of the detrended series
values were identical, but not the tails, as the two routines do not use
the same data to establish the tails. These differences produce
different GSI (\texttt{dfoliatR}) and Corrected Indices (OUTBREAK)
values at the tails, sometimes considerably, affecting whether or not
outbreak events are inferred early or late in the tree series, timing of
the first or last event onset, and subsequent computation of return
intervals, duration, and periodicity. Note that there is nothing
particularly important regarding the choice of ARSTAN 6.1 -- it is
simply the version that the second author used routinely and had
available.

@@

\texttt{dfoliatR} and OUTBREAK compute growth suppresses indices and use
them to infer identical tree-level outbreak events, including onset and
termination dates and date and magnitude of the maximum growth
suppression when \texttt{dfoliatR} bridging matches the OUTBREAK
protocols. Once the issues with ARSTAN (described earlier) were
resolved, dfoliatR and OUTBREAK produced identical indices at 0.000
precision.

When inferring outbreak events at the tree level, our initial results
did not always match OUTBREAK, leading to development of the bridging
option. Identifying a minimum within a period while allowing for
positive excursions is a tricky computation. OUTBREAK deliberately
disallows back-to-back events -- after a minimum value within a period
of negative values, a second positive index terminates an event.
OUTBREAK does not allow a period of highly negative values to be
appended to an earlier outbreak if the intervening period is 1 year,
even if those negative values fall below the minimum in the earlier
period. Due in large part to reconstructions using OUTBREAK (see papers
cited by Lynch 2012), we now know considerably more about forest
defoliator outbreak regimes than we did in the 1980s when OUTBREAK was
written. We think that two prolonged events separated by a single year
should in some situations be considered a single outbreak at both the
tree and site levels. This is particularly relevant to western spruce
budworm and spruce budworm (\emph{C. fumiferana}), for which regimes
have been reconstructed for several geographic areas and for which
researchers and forest health experts have gained considerable knowledge
(\textbf{Sanders et al.~1985, Brookes et al.~1987 and many later
publications}). We now know that the greatest growth suppression often
occurs late in the outbreak due to cumulative effects (\textbf{rrr}). We
developed the bridging option to permit these periods to be linked to
preceding events separated by a single year.

Bridging operates at the tree level in dfoliatR, and these effects
generally have minor effects on site-level reconstructions
(Fig.Bridging). Occasionally site-level events are inferred differently
(Table.Bridging), and this can affect return interval and duration
statistics.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={dfoliatR: An R package for detection and analysis of insect defoliation signals in tree rings},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage{amsmath}
\usepackage{float}
\usepackage{booktabs}

@@

Variation in the width and morphology of annual radial growth rings in trees permits dating and quantification of past forest insect defoliator outbreaks. Defoliation can be distinguished from climate- and other disturbance-related influences by comparing ring-width or other annually-resolved features in the wood of host species to non-host species or to climate records. The effect of defoliation on radial growth of trees has been recognized since the mid-1800s, but it was not until the 1980s that dendrochronology techniques were applied to identify and quantify defoliation events (Lynch, 2012; Speer, 2010; Swetnam et al., 1985). The first studies (Swetnam and Lynch, 1993, 1989; Swetnam et al., 1985) focused on developing historical reconstructions of western spruce budworm (WSBW; \emph{Choristoneura freemani}). The methodology has since been successfully applied to a wide range of defoliator species, most of which are conifer herbivores, and has evolved in sophistication and application to a wide range of ecosystem situations (Lynch, 2012).

The main dendrochronological tool for inferring, dating, and quantifying defoliator outbreaks from tree-ring records has been the software routine OUTBREAK (Holmes and Swetnam, 1986; Swetnam and Lynch, 1989; Swetnam et al., 1985). OUTBREAK computes indices (described later in detail) of suppressed growth by subtracting a detrended and standardized climate series (a ``control'' chronology) from host individual-tree detrended and standardized radial growth series after the host and non-host series have been brought to a common variance. The non-host chronology often consisted of a site chronology developed from non-host tree species, but a gridded climate data point series, like the North American Drought Atlas (Cook and Krusic, 2004) also suffices. If the host and non-host species respond similarly to climate (which can be tested), the derived series retains variability that the host and non-host series do not have in common, generally the insect signal, along with some unexplained variability (noise). The user defines a rule base specifying the magnitude and duration that a period of indexed growth suppression must meet or surpass for a period of suppressed growth to be inferred as a defoliation event at the tree level.

Though powerful, OUTBREAK is outdated and increasingly difficult to use in modern computing environments. It was written in FORTRAN V with inherently severe restrictions, as RAM and disk space were limited at that time (256 kb and 10 MB, respectively) and FORTRAN conventions imposed very strict formatting, file naming, and output conventions. The program provides no graphical interface or capabilities, forcing users to import generated text files into spreadsheet or other software for assessing results and performing analyses. Furthermore, OUTBREAK can only handle one test at a time, creating barriers to batch operation and a large burden for researchers with a datasets including multiple sites. We developed the software library \texttt{dfoliaR} as an \texttt{R}- and \texttt{dplR}-based routine to overcome these issues.

\texttt{dfoliatR} adds to a growing suite of dendrochronology packages the \texttt{R} computing environment (R Core Team, 2019). Stemming from the \texttt{dplR} library (Bunn, 2008) that enables \texttt{R} users to read and write an array of tree-ring data formats, standardize ring width series, build and evaluate chronologies, and perform quality control (to name a few), one can now also measure ring widths from scanned images of prepared samples (Lara et al., 2015; Shi and Xiang, 2019), perform and check crossdating (Bunn, 2010), and perform many analytical tests (Jevšenak and Levanič, 2018; Zang and Biondi, 2015). Tools for assessing stand dynamics and disturbance analyses are under rapid development, with new packages for assessing growth and release events (\texttt{TRADER}: Altman et al., 2014), metrics of growth resilience (\texttt{pointRes}: Maaten-Theunissen et al., 2015), and fire history (\texttt{burnr}: Malevich et al., 2018). The key objective of \texttt{dfoliatR} is to provide tools to identify and analyze insect defoliation and outbreak events by building on the methods employed by OUTBREAK. It capitalizes on the robust software already available in \texttt{R} by using \texttt{dplR} data formats for incoming tree-ring series and providing output data formats embodied by the \texttt{tidyverse} (Wickham et al., 2019) that include efficient data manipulation (Wickham et al., 2020) and graphics (Wickham, 2016).

In this paper, we describe the statistical methods employed by \texttt{dfoliatR}, compare results to those produced by OUTBREAK, and present an example analysis. Users need not have much experience in \texttt{R} to replicate the analyses and graphics as presented. The \texttt{R} code below is executable in an \texttt{R} session once the required libraries are installed and loaded. Support documentation in addition to this paper is provided within the package via standard help menus (accessed by typing \texttt{?} before a function name) and on the package website (\url{https://chguiterman.github.io/dfoliatR/}), which includes up-to-date vignettes that describe various software routines. Code to generate a preprint of this manuscript including the \texttt{R} scripts, and tubular and graphical output is available from \url{https://github.com/chguiterman/dfoliatR_paper}.

@@

Users can develop these data sets in software of their choosing, such as \texttt{dplR} (Bunn, 2008) or ARSTAN (Cook and Holmes, 1996). It is important that the host-tree data include only one tree-ring series per tree. Both \texttt{dplR} and ARSTAN have options for averaging multiple sample series into a tree-level series.

At the heart of \texttt{dfoliatR} lies two functions: \texttt{defoliate\_trees()} and \texttt{outbreak()}. These identify defoliation events on individual trees (Figure \ref{fig:fig-defol}) and then composite across multiple trees to identify outbreak events (Figure \ref{fig:fig-obr}).

@@

The \texttt{defoliate\_trees()} function is the point of entry to the \texttt{dfoliatR} library. It performs two processes, removing climate-related growth signals from the host-tree series and identifying tree-level defoliation events. The climatic or non-defoliation signals in each host-tree series are characterized by a non-host chronology or climate reconstruction. \texttt{dfoiatR} removes that non-defoliation signal by subtracting the non-host series from each host-tree series, which generates a residual index. In OUTBREAK, this residual index was termed the ``corrected'' index. We call it the ``growth suppression index'' (GSI). The GSI is calculated the same as in OUTBREAK for each host tree as
\begin{align}
\textrm{GSI}_{i} = \textrm{H}_{i} - \left( \textrm{NH}_{i} - \overline{\textrm{NH}} \right) \frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}} \
\end{align}
where H and NH are the host-tree series and the non-host chronology, in year i, respectively. Only the common period between the host-tree series and the non-host chronology are used in Equation 1. The host and non-host chronologies are brought to common variance by scaling the non-host chronology by its mean (\(\overline{\textrm{NH}} \approx 1.0\)) and multiplying by the ratio of host and non-host standard deviations (\(\frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}}\)), which approximates the variance of the host tree series.

Negative departures in the normalized GSI (NGSI) that surpass user-defined thresholds in duration and magnitude are defined as \emph{defoliation events}. As in OUTBREAK, magnitude is assessed on a single year within the departure sequence. The default setting is -1.28 (in units of standard deviation), which was previously determined to be representative of WSBW effects (Swetnam and Lynch, 1989). Duration is assessed by examining sequences of negative NGSI before and after the year of maximum departure. Each defoliation event is allowed one positive excursion on each side of the maximum departure year. Duration is assessed across the entire sequence that includes up to two positive excursions. The default duration is eight years, as is commonly used in WSBW studies (Swetnam and Lynch, 1989). Different species of defoliation insects vary in the length of defoliation and the degree to which they can suppress tree growth. Researchers can, and should, adjust the duration and magnitude parameters accordingly and critically evaluate the results.

@@

Defoliation of one or a few trees does not constitute an outbreak. To determine when defoliation becomes an \emph{outbreak event}, \texttt{dfoliatR} composites the individual tree defoliation series into a site-level chronology with the \texttt{outbreak()} function. Users have options to define the number and/or the proportion of trees required for an event to be considered an outbreak. Three parameters control whether a defoliation event constitutes an outbreak: the minimum number of trees available, the minimum number of trees recording defoliation, and the percent of trees recording defoliation. The first allows the researcher to make a judgement call as to the confidence ascribed to reduced sample depth toward the ends of their chronologies, thus compensating for the ``fading record problem'' (Swetnam et al., 1999). The second two parameters adjust the scale of defoliation considered to be an outbreak. Absolute numbers of trees and percentages can be applied separately or in conjunction, following filtering conventions in tree ring fire history studies (Malevich et al., 2018). We urge users to carefully consider the choice of absolute numbers in situations where the number of trees represented in the series varies with time, or the choice of percentages when sample size is small.

@@

We tested \texttt{dfoliatR} against OUTBREAK by comparing NGSI to corrected indices for individual trees and years, defoliation status for individual trees and years, and percentage of trees recording outbreaks at the site level, using standardized ring-width data from eight sites in British Columbia, Colorado, and New Mexico, and author-provided non-host site chronologies.

We detrended host data for both \texttt{dfoliatR} and OUTBREAK using ARSTAN v6.1 (Cook and Holmes, 1996) with default double detrending (128 year wavelength and a 50\% smoothing spline). In both \texttt{dfoliatR} and OUTBREAK we used event thresholds of -1.28 normalized index, 8 years duration, and allowed for events at the end of series in seven of eight sites. These seven sites were sampled during ongoing outbreak events (Swetnam and Lynch, 1993). We found it necessary to be consistent in how we detrended and in what software (e.g., ARSTAN vs \texttt{dplR}) we employed because subtle differences in standardized ring-width indices created between the programs exacerbated differences between \texttt{dfoliatR} and OUTBREAK. In the end, we chose to only use the standardization output files from ARSTAN, which are easily read into \texttt{R} (and then \texttt{dfoliatR}) using the \texttt{dplR} package.

The \texttt{R} code to replicate our comparisons is available from \url{https://github.com/chguiterman/dfoliatR_paper}.

@@

Since both programs employ Equation 1, \texttt{dfoliatR} and OUTBREAK compute identical growth suppression indices at 0.000 precision (Figure \ref{fig:df-obr-1-1}). In all there are 43,280 ring-width indices from 222 trees included in our test. Of these, the programs identified 11,530 years of defoliation. The programs agreed on 92\% of the years, leaving 927 ``difference'' years in which only one program identified defoliation on an individual tree. The differences included 102 events on 85 trees. We carefully inspected each of these events in the full context of each tree's ring series, and categorized the differences as follows

\begin{itemize}
\item
  \emph{Series-end events} (40\% of the total) in which we allowed OUTBREAK to include ``truncated outbreaks'' (for seven sites) at the end of each series. In \texttt{dfoliatR}, this option is controlled by the ``series\_end\_events'' parameter to \texttt{defoliate\_trees()}. In OUTBREAK, the option appears while changing the duration parameter (option 3). When selected, OUTBREAK will include any sequences of negative indices at the either the beginning or the end of each tree series. No thresholds are applied in identifying these as defoliation events. In \texttt{dfoliatR}, the negative sequences are evaluated as with any other possible defoliation, but the duration threshold is omitted. Each of the 13 events included in these differences did not meet the ``max\_reduction'' parameter (-1.28 NGSI) in \texttt{dfoliatR} and were excluded. In two cases, OUTBREAK included events at the beginning of the series where \texttt{dfoliatR} does not allow truncated events. In four cases, OUTBREAK omitted the last year of the series because it was positive. In two cases, \texttt{dfoliatR} omitted possible events because it had already included a positive NGSI excursion after the ``max\_reduction'' year, and since it will only allow one excursion on either side of the max year, the events were omitted due to short duration.
\item
  \emph{Sequential events} (36\%) in which OUTBREAK omitted events that occur one year prior or one year following an identified event. OUTBREAK selects the event with the greatest negative index year. On two trees, OUTBREAK omitted two of three sequential events. \texttt{dfoliatR} allows all such events. While inspecting these differences, we added an option to \texttt{defoiate\_trees()} that would ``bridge'' between sequential events, making them into single, long events. We felt that this was ecologically justified, especially for studies of WSBW, because the insect will occasionally maintain minor defoliation f individuals for decades even while not in an outbreak phase.
\item
  \emph{Undetermined differences} (22\%) occurred in cases where OUTBREAK omitted events without clear cause.
\item
  \emph{Rounding} (2\%) differences in the indices either omitted or cut short events on two trees. In both cases the indices were very close to zero.
\end{itemize}

Our inspections revealed what we believe are short-comings in how OUTBREAK identifies defoliation events on individual trees. In every one of the 102 cases we inspected, we felt that \texttt{dfoliatR} provided a more biologically and statistically appropriate assessment of defoliation.

and use them to infer identical tree-level outbreak events, including onset and termination dates and date and magnitude of the maximum growth suppression when \texttt{dfoliatR} bridging matches the OUTBREAK protocols. Once the issues with ARSTAN (described earlier) were resolved, dfoliatR and OUTBREAK produced identical indices at 0.000 precision.
